---
title: "Minimal SFC-IO Tutorial"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

This `.qmd` is the single source of truth.

- Generate standalone script: `Rscript build_tutorial.R --purl-only`
- Render notebook html: `Rscript build_tutorial.R`
- Render notebook + slides: `Rscript build_tutorial.R --reveal`

## Setup and Data

```{r helpers}
as_numeric_clean <- function(x) {
  y <- gsub(",", "", as.character(x), fixed = TRUE)
  y <- gsub(" ", "", y, fixed = TRUE)
  y <- gsub("[^0-9eE+\\-\\.]", "", y)
  suppressWarnings(as.numeric(y))
}

safe_div <- function(num, den, default = 0) {
  out <- rep(default, length(num))
  ok <- is.finite(num) & is.finite(den) & den != 0
  out[ok] <- num[ok] / den[ok]
  out
}

check_setup <- function() {
  required_packages <- c("readxl", "ggplot2")
  missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing_packages) > 0) {
    stop("Missing package(s): ", paste(missing_packages, collapse = ", "), call. = FALSE)
  }

  required_files <- c("IOT_AT_2021_TOTAL_energy_green_brown.xlsx", "wealth_assets.xlsx")
  missing_files <- required_files[!file.exists(required_files)]
  if (length(missing_files) > 0) {
    stop("Missing data file(s): ", paste(missing_files, collapse = ", "), call. = FALSE)
  }
  invisible(TRUE)
}

load_data <- function(base_year = 2021) {
  check_setup()

  io <- as.data.frame(readxl::read_xlsx(
    "IOT_AT_2021_TOTAL_energy_green_brown.xlsx",
    sheet = "Sheet 1",
    range = "A11:CE95",
    col_names = TRUE
  ))

  wealth <- readxl::read_xlsx(
    "wealth_assets.xlsx",
    sheet = "nasa_10_f_bs",
    range = "A10:M19",
    col_names = TRUE
  )

  label_col <- "PRD_USE (Labels)"
  if (!(label_col %in% names(io))) stop("Missing PRD_USE labels in IO.", call. = FALSE)

  sector_labels <- as.character(io[[label_col]][2:66])
  if (length(sector_labels) != 65) stop("Expected 65 sectors.", call. = FALSE)

  co2_sheet <- tail(readxl::excel_sheets("IOT_AT_2021_TOTAL_energy_green_brown.xlsx"), 1)
  co2_tbl <- readxl::read_xlsx(
    "IOT_AT_2021_TOTAL_energy_green_brown.xlsx",
    sheet = co2_sheet,
    col_names = TRUE
  )
  co2_match <- match(sector_labels, as.character(co2_tbl[["To Sector"]]))
  if (any(is.na(co2_match))) stop("Missing CO2 intensities for some sectors.", call. = FALSE)

  list(
    base_year = base_year,
    io = io,
    wealth = wealth,
    sector_labels = sector_labels,
    co2_tbl = co2_tbl,
    co2_match = co2_match
  )
}

calibrate_base <- function(data) {
  io <- data$io
  wealth <- data$wealth
  n <- 65

  Z <- matrix(as_numeric_clean(unlist(io[2:66, 3:67])), nrow = n, ncol = n)
  q_s <- as_numeric_clean(io[84, 3:67])
  q_s[!is.finite(q_s) | q_s <= 0] <- 1e-9

  A <- sweep(Z, 2, q_s, "/")
  A[!is.finite(A)] <- 0
  L <- solve(diag(n) - A)

  C_i0 <- as_numeric_clean(io[2:66, "Final consumption expenditure by households"]) +
    as_numeric_clean(io[2:66, "Final consumption expenditure by non-profit organisations serving households (NPISH)"])
  G_i0 <- as_numeric_clean(io[2:66, "Final consumption expenditure by government"])
  I_i0 <- as_numeric_clean(io[2:66, "Gross capital formation...72"])
  EX_i0 <- as_numeric_clean(io[2:66, "Exports of goods and services"])
  X_i0 <- as_numeric_clean(io[2:66, "Total use"])

  C0 <- sum(C_i0)
  G0 <- sum(G_i0)

  beta_C <- safe_div(C_i0, rep(C0, n))
  beta_G <- safe_div(G_i0, rep(G0, n))
  beta_C <- beta_C / sum(beta_C)
  beta_G <- beta_G / sum(beta_G)

  imports_row <- which(io[["PRD_USE (Labels)"]] == "Imports of goods and services")
  imports_i <- as_numeric_clean(io[imports_row, 3:67])
  m_i <- pmin(pmax(safe_div(imports_i, X_i0), 0), 0.95)

  va_row <- which(io[["PRD_USE (Labels)"]] == "Value added, gross")
  va_i <- as_numeric_clean(io[va_row, 3:67])
  va_coeff <- safe_div(va_i, q_s)
  va_coeff[!is.finite(va_coeff) | va_coeff < 0] <- 0

  Y0 <- sum(va_coeff * X_i0)
  tau_y <- (as_numeric_clean(io[73, "Total use"]) + as_numeric_clean(io[74, "Total use"])) / Y0
  tau_y <- ifelse(is.finite(tau_y), pmin(pmax(tau_y, 0), 0.6), 0.2)

  w_assets_col <- names(wealth)[2]
  w_liab_col <- names(wealth)[3]
  hh_row <- which(wealth[["NA_ITEM (Labels)"]] == "Households")
  npish_row <- which(grepl("^Non-profit institutions", wealth[["NA_ITEM (Labels)"]]))
  gov_row <- which(wealth[["NA_ITEM (Labels)"]] == "General government")

  hh_assets <- as_numeric_clean(wealth[hh_row, w_assets_col])
  hh_liab <- as_numeric_clean(wealth[hh_row, w_liab_col])
  npish_assets <- if (length(npish_row) == 1) as_numeric_clean(wealth[npish_row, w_assets_col]) else 0
  npish_liab <- if (length(npish_row) == 1) as_numeric_clean(wealth[npish_row, w_liab_col]) else 0

  V0 <- (hh_assets - hh_liab) + (npish_assets - npish_liab)
  B0 <- as_numeric_clean(wealth[gov_row, w_liab_col])

  s_prod <- as_numeric_clean(data$co2_tbl[["Mean CO2 intensity (kg/M€)"]][data$co2_match])
  s_prod[!is.finite(s_prod)] <- 0

  list(
    base_year = data$base_year,
    n = n,
    sector_labels = data$sector_labels,
    L = L,
    va_coeff = va_coeff,
    beta_C = beta_C,
    beta_G = beta_G,
    m_i = m_i,
    I_i0 = I_i0,
    EX_i0 = EX_i0,
    tau_y = tau_y,
    V0 = V0,
    B0 = B0,
    G0 = G0,
    s_prod = s_prod,
    idx_brown = match("Energy Brown", data$sector_labels),
    idx_green = match("Energy Green", data$sector_labels)
  )
}
```

```{r load-and-calibrate}
check_setup()
data_obj <- load_data(2021)
calib <- calibrate_base(data_obj)

data.frame(
  metric = c("Sectors", "tau_y", "G0", "V0", "B0"),
  value = c(calib$n, calib$tau_y, calib$G0, calib$V0, calib$B0)
)
```

## Step 1: Simplest Model (Closed SFC-IO)

```{r step1-function}
simulate_step1_closed <- function(calib, T = 20, c_y = 0.85, c_v = 0.02, g_G = 0.01) {
  n <- calib$n
  years <- calib$base_year + seq_len(T) - 1

  out <- data.frame(
    year = years,
    GDP = numeric(T), TAX = numeric(T), YD = numeric(T),
    C = numeric(T), G = numeric(T), V = numeric(T),
    DEF = numeric(T), B = numeric(T)
  )
  x_mat <- matrix(0, nrow = T, ncol = n)

  V_prev <- calib$V0
  B_prev <- calib$B0
  G_prev <- calib$G0

  k <- as.numeric(t(calib$va_coeff) %*% (calib$L %*% calib$beta_C))

  for (t in seq_len(T)) {
    G_t <- if (t == 1) calib$G0 else G_prev * (1 + g_G)

    other_fd <- calib$beta_G * G_t + calib$I_i0
    b <- as.numeric(t(calib$va_coeff) %*% (calib$L %*% other_fd))

    den <- 1 - c_y * (1 - calib$tau_y) * k
    C_t <- (c_y * (1 - calib$tau_y) * b + c_v * V_prev) / den
    C_t <- max(C_t, 0)

    fd_i <- calib$beta_C * C_t + other_fd
    x_t <- as.numeric(calib$L %*% fd_i)
    Y_t <- sum(calib$va_coeff * x_t)

    TAX_t <- calib$tau_y * Y_t
    YD_t <- Y_t - TAX_t

    V_t <- V_prev + YD_t - C_t
    DEF_t <- G_t - TAX_t
    B_t <- B_prev + DEF_t

    out[t, c("GDP", "TAX", "YD", "C", "G", "V", "DEF", "B")] <- c(Y_t, TAX_t, YD_t, C_t, G_t, V_t, DEF_t, B_t)
    x_mat[t, ] <- x_t

    V_prev <- V_t
    B_prev <- B_t
    G_prev <- G_t
  }

  list(aggregate = out, x = x_mat)
}
```

```{r step1-run}
step1 <- simulate_step1_closed(calib, T = 20, c_y = 0.85, c_v = 0.02, g_G = 0.01)
head(step1$aggregate)
```

```{r step1-plot, fig.height=4, fig.width=7}
ggplot2::ggplot(step1$aggregate, ggplot2::aes(year, GDP)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(title = "Step 1: GDP (closed model)", y = "M EUR") +
  ggplot2::theme_minimal()
```

## Step 2: Add RoW (Imports and Exports)

```{r step2-function}
simulate_step2_open <- function(calib, T = 20, c_y = 0.85, c_v = 0.02, g_G = 0.01, ex_growth = 0) {
  n <- calib$n
  years <- calib$base_year + seq_len(T) - 1
  S <- 1 - calib$m_i

  out <- data.frame(
    year = years,
    GDP = numeric(T), TAX = numeric(T), YD = numeric(T),
    C = numeric(T), G = numeric(T), V = numeric(T),
    DEF = numeric(T), B = numeric(T),
    EX = numeric(T), IM = numeric(T), TB = numeric(T)
  )
  x_mat <- matrix(0, nrow = T, ncol = n)

  V_prev <- calib$V0
  B_prev <- calib$B0
  G_prev <- calib$G0

  k <- as.numeric(t(calib$va_coeff) %*% (calib$L %*% (S * calib$beta_C)))

  for (t in seq_len(T)) {
    G_t <- if (t == 1) calib$G0 else G_prev * (1 + g_G)
    EX_i_t <- calib$EX_i0 * (1 + ex_growth)^(t - 1)

    other_fd <- calib$beta_G * G_t + calib$I_i0 + EX_i_t
    b <- as.numeric(t(calib$va_coeff) %*% (calib$L %*% (S * other_fd)))

    den <- 1 - c_y * (1 - calib$tau_y) * k
    C_t <- (c_y * (1 - calib$tau_y) * b + c_v * V_prev) / den
    C_t <- max(C_t, 0)

    fd_dom_i <- calib$beta_C * C_t + other_fd
    im_i <- calib$m_i * fd_dom_i
    fd_net_i <- pmax(fd_dom_i - im_i, 0)

    x_t <- as.numeric(calib$L %*% fd_net_i)
    Y_t <- sum(calib$va_coeff * x_t)

    TAX_t <- calib$tau_y * Y_t
    YD_t <- Y_t - TAX_t

    V_t <- V_prev + YD_t - C_t
    DEF_t <- G_t - TAX_t
    B_t <- B_prev + DEF_t

    EX_t <- sum(EX_i_t)
    IM_t <- sum(im_i)
    TB_t <- EX_t - IM_t

    out[t, c("GDP", "TAX", "YD", "C", "G", "V", "DEF", "B", "EX", "IM", "TB")] <-
      c(Y_t, TAX_t, YD_t, C_t, G_t, V_t, DEF_t, B_t, EX_t, IM_t, TB_t)

    x_mat[t, ] <- x_t

    V_prev <- V_t
    B_prev <- B_t
    G_prev <- G_t
  }

  list(aggregate = out, x = x_mat)
}

run_step2_scenarios <- function(calib, T = 20) {
  specs <- list(
    baseline = list(g_G = 0.01, c_y = 0.85, c_v = 0.02, ex_growth = 0),
    fiscal_expansion = list(g_G = 0.03, c_y = 0.85, c_v = 0.02, ex_growth = 0),
    high_wealth_consumption = list(g_G = 0.01, c_y = 0.85, c_v = 0.06, ex_growth = 0)
  )

  res <- lapply(names(specs), function(nm) {
    sp <- specs[[nm]]
    sim <- simulate_step2_open(calib, T = T, c_y = sp$c_y, c_v = sp$c_v, g_G = sp$g_G, ex_growth = sp$ex_growth)
    sim$aggregate$scenario <- nm
    sim
  })
  names(res) <- names(specs)

  agg <- do.call(rbind, lapply(res, function(x) x$aggregate))
  rownames(agg) <- NULL
  list(by_scenario = res, aggregate = agg)
}
```

```{r step2-run}
step2 <- run_step2_scenarios(calib, T = 20)
head(step2$aggregate)
```

```{r step2-plots, fig.height=4, fig.width=7}
p1 <- ggplot2::ggplot(step2$aggregate, ggplot2::aes(year, GDP, color = scenario)) +
  ggplot2::geom_line(linewidth = 1) + ggplot2::theme_minimal() +
  ggplot2::labs(title = "Step 2: GDP with RoW")

p2 <- ggplot2::ggplot(step2$aggregate, ggplot2::aes(year, TB, color = scenario)) +
  ggplot2::geom_line(linewidth = 1) + ggplot2::theme_minimal() +
  ggplot2::labs(title = "Step 2: Trade balance")

print(p1)
print(p2)
```

## Step 3: Steady-State Play (scan `c_v`)

```{r step3-function}
steady_state_play <- function(calib, cv_grid = seq(0, 0.12, by = 0.01), target_window = 5) {
  out <- data.frame(c_v = cv_grid, avg_gdp_growth_last_window = NA_real_)

  for (i in seq_along(cv_grid)) {
    sim <- simulate_step2_open(calib, T = 20, c_y = 0.85, c_v = cv_grid[i], g_G = 0.01, ex_growth = 0)
    gdp <- sim$aggregate$GDP
    growth <- c(NA_real_, gdp[-1] / gdp[-length(gdp)] - 1)
    out$avg_gdp_growth_last_window[i] <- mean(tail(growth[is.finite(growth)], target_window))
  }

  best <- out[which.min(abs(out$avg_gdp_growth_last_window)), , drop = FALSE]
  list(grid = out, best = best)
}
```

```{r step3-run}
step3 <- steady_state_play(calib, cv_grid = seq(0, 0.12, by = 0.01), target_window = 5)
step3$grid
```

```{r step3-best}
step3$best
```

## Step 4: Add CO2 (from sector output)

```{r step4-function}
add_co2 <- function(sim_out, calib) {
  co2 <- rowSums(sim_out$x * matrix(rep(calib$s_prod, each = nrow(sim_out$x)), nrow = nrow(sim_out$x))) / 1e9
  out <- sim_out$aggregate
  out$CO2_Mt <- co2
  out
}

add_co2_to_step2 <- function(step2_out, calib) {
  agg_list <- lapply(step2_out$by_scenario, function(sim) {
    out <- add_co2(sim, calib)
    out$scenario <- unique(sim$aggregate$scenario)
    out
  })
  agg <- do.call(rbind, agg_list)
  rownames(agg) <- NULL
  agg
}
```

```{r step4-run}
step4_agg <- add_co2_to_step2(step2, calib)
head(step4_agg)
```

```{r step4-plot, fig.height=4, fig.width=7}
ggplot2::ggplot(step4_agg, ggplot2::aes(year, CO2_Mt, color = scenario)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(title = "Step 4: CO2 emissions", y = "Mt CO2") +
  ggplot2::theme_minimal()
```

## Step 5: Add Transition (shift brown to green demand)

```{r step5-function}
simulate_step5_transition <- function(calib, T = 20, c_y = 0.85, c_v = 0.02, g_G = 0.01,
                                      ex_growth = 0, annual_shift = 0.002, enabled = TRUE) {
  n <- calib$n
  years <- calib$base_year + seq_len(T) - 1
  S <- 1 - calib$m_i

  out <- data.frame(
    year = years,
    GDP = numeric(T), C = numeric(T), G = numeric(T),
    V = numeric(T), B = numeric(T), TB = numeric(T), CO2_Mt = numeric(T),
    betaC_brown = numeric(T), betaC_green = numeric(T)
  )

  V_prev <- calib$V0
  B_prev <- calib$B0
  G_prev <- calib$G0

  beta_C0 <- calib$beta_C
  beta_G0 <- calib$beta_G

  x_mat <- matrix(0, nrow = T, ncol = n)

  for (t in seq_len(T)) {
    G_t <- if (t == 1) calib$G0 else G_prev * (1 + g_G)
    EX_i_t <- calib$EX_i0 * (1 + ex_growth)^(t - 1)

    beta_C_t <- beta_C0
    beta_G_t <- beta_G0

    if (enabled && is.finite(calib$idx_brown) && is.finite(calib$idx_green)) {
      shift <- (t - 1) * annual_shift
      shift_C <- min(shift, max(0, beta_C0[calib$idx_brown] - 1e-8))
      shift_G <- min(shift, max(0, beta_G0[calib$idx_brown] - 1e-8))

      beta_C_t[calib$idx_brown] <- beta_C0[calib$idx_brown] - shift_C
      beta_C_t[calib$idx_green] <- beta_C0[calib$idx_green] + shift_C

      beta_G_t[calib$idx_brown] <- beta_G0[calib$idx_brown] - shift_G
      beta_G_t[calib$idx_green] <- beta_G0[calib$idx_green] + shift_G
    }

    k <- as.numeric(t(calib$va_coeff) %*% (calib$L %*% (S * beta_C_t)))
    other_fd <- beta_G_t * G_t + calib$I_i0 + EX_i_t
    b <- as.numeric(t(calib$va_coeff) %*% (calib$L %*% (S * other_fd)))

    den <- 1 - c_y * (1 - calib$tau_y) * k
    C_t <- (c_y * (1 - calib$tau_y) * b + c_v * V_prev) / den
    C_t <- max(C_t, 0)

    fd_dom_i <- beta_C_t * C_t + other_fd
    im_i <- calib$m_i * fd_dom_i
    fd_net_i <- pmax(fd_dom_i - im_i, 0)

    x_t <- as.numeric(calib$L %*% fd_net_i)
    Y_t <- sum(calib$va_coeff * x_t)
    T_t <- calib$tau_y * Y_t
    YD_t <- Y_t - T_t

    V_t <- V_prev + YD_t - C_t
    B_t <- B_prev + (G_t - T_t)

    TB_t <- sum(EX_i_t) - sum(im_i)
    CO2_t <- sum(calib$s_prod * x_t) / 1e9

    out[t, c("GDP", "C", "G", "V", "B", "TB", "CO2_Mt", "betaC_brown", "betaC_green")] <-
      c(Y_t, C_t, G_t, V_t, B_t, TB_t, CO2_t,
        if (is.finite(calib$idx_brown)) beta_C_t[calib$idx_brown] else NA_real_,
        if (is.finite(calib$idx_green)) beta_C_t[calib$idx_green] else NA_real_)

    x_mat[t, ] <- x_t

    V_prev <- V_t
    B_prev <- B_t
    G_prev <- G_t
  }

  list(aggregate = out, x = x_mat)
}
```

```{r step5-run}
baseline_open <- simulate_step5_transition(calib, T = 20, enabled = FALSE)
green_transition <- simulate_step5_transition(calib, T = 20, enabled = TRUE, annual_shift = 0.002)

compare <- data.frame(
  year = baseline_open$aggregate$year,
  CO2_baseline = baseline_open$aggregate$CO2_Mt,
  CO2_transition = green_transition$aggregate$CO2_Mt,
  brown_share_baseline = baseline_open$aggregate$betaC_brown,
  brown_share_transition = green_transition$aggregate$betaC_brown
)

head(compare)
```

```{r step5-plot, fig.height=4, fig.width=7}
ggplot2::ggplot(compare, ggplot2::aes(year)) +
  ggplot2::geom_line(ggplot2::aes(y = CO2_baseline, color = "baseline"), linewidth = 1) +
  ggplot2::geom_line(ggplot2::aes(y = CO2_transition, color = "transition"), linewidth = 1) +
  ggplot2::labs(title = "Step 5: CO2 baseline vs transition", y = "Mt CO2", color = "Scenario") +
  ggplot2::theme_minimal()
```

## Final Snapshot

```{r final-snapshot}
last_year <- max(step4_agg$year)
summary_final <- step4_agg[step4_agg$year == last_year, c("scenario", "GDP", "V", "B", "TB", "CO2_Mt")]
summary_final <- summary_final[order(summary_final$scenario), ]
rownames(summary_final) <- NULL
summary_final
```
